% ============================================================
% ML Defender (aRGus NDR) — arXiv preprint
% Draft v5 — DAY 88 — March 2026
% Author: Alonso Isidoro Roman
% Target: cs.CR
% ============================================================

\documentclass[11pt,a4paper]{article}

% --- Encoding and fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% --- Language ---
\usepackage[english]{babel}

% --- Math ---
\usepackage{amsmath}
\usepackage{amssymb}

% --- Layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing

% --- Tables ---
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{array}

% --- Code listings ---
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!10},
  frame=single,
  rulecolor=\color{gray!40},
  breaklines=true,
  breakatwhitespace=true,
  columns=flexible,
  keepspaces=true,
  showstringspaces=false
}

% --- Hyperlinks ---
\usepackage[colorlinks=true, linkcolor=blue!70!black,
            citecolor=blue!70!black, urlcolor=blue!70!black]{hyperref}

% --- Graphics ---
\usepackage{graphicx}

% --- References ---
\usepackage[numbers,sort&compress]{natbib}

% --- Misc ---
\usepackage{xspace}
\usepackage{enumitem}

% ============================================================
% Title, author, date
% ============================================================

\title{\textbf{ML Defender (aRGus NDR): An Open-Source Embedded ML NIDS\\
for Botnet and Anomalous Traffic Detection\\
in Resource-Constrained Organizations}}

\author{
  Alonso Isidoro Rom\'an\\
  \textit{Independent Researcher, Extremadura, Spain}\\
  \textit{Universidad de Extremadura (informal collaboration)}\\
  \href{https://github.com/alonsoir/argus}{github.com/alonsoir/argus}
}

\date{Draft v5 --- March 2026}

% ============================================================
\begin{document}
% ============================================================

\maketitle

\begin{abstract}
Ransomware and DDoS attacks disproportionately impact hospitals, schools, and small
organizations that cannot afford enterprise security solutions. Existing open-source
alternatives either rely on signature matching --- missing novel variants --- or require
infrastructure beyond the reach of their target users. We present \textbf{ML Defender
(aRGus NDR)}, an open-source network intrusion detection system built in C++20 that brings
embedded machine learning inference to organizations with limited resources, deployable on
commodity bare-metal hardware at approximately 150--200~USD fully dedicated.

ML Defender implements a six-component pipeline over eBPF/XDP packet capture, ZeroMQ
transport, and Protocol Buffers serialization, terminating in a dual-score detection
architecture combining a rule-based Fast Detector with an embedded Random Forest classifier.
The \emph{Maximum Threat Wins} policy selects the arithmetic maximum of both detector scores,
using ML inference to suppress false positives generated by the heuristic layer.

We evaluate ML Defender against the CTU-13 Neris botnet dataset, achieving
$\text{F1} = 0.9985$, $\text{Precision} = 0.9969$, and $\text{Recall} = 1.0000$,
with a false positive rate of 0.0002\% (2~FP in 12,075 benign flows) --- the two false
positives being VirtualBox network artifacts (multicast/broadcast) that do not occur in
bare-metal deployments. The Fast Detector alone produces a FPR of 6.61\% on purely benign
traffic (bigFlows); the ML Detector reduces real production blocks to zero on the same
corpus, an approximately 500-fold reduction. Per-class inference latency ranges from
0.24~$\mu$s (DDoS) to 1.06~$\mu$s (Ransomware), enabling real-time detection on commodity
hardware.

We further characterize the pipeline's throughput ceiling under progressive load in a
virtualized environment. Requesting up to 100~Mbps via \texttt{tcpreplay} against the
VirtualBox NIC, the pipeline processes up to ${\sim}34$--38~Mbps (the emulated NIC ceiling)
with zero packet drops and zero errors across 2.37 million packets. At peak load,
\texttt{ml-detector} consumes approximately 3.2 CPU cores; total pipeline RAM remains stable
at ${\sim}1.28$~GB with negligible drift (${\sim}18$~MB over 8 minutes). The bottleneck is
VirtualBox NIC emulation, not pipeline logic. Post-replay drain behavior confirms that
ZeroMQ queues absorb burst traffic correctly and drain asynchronously, validating the queue
stability design. All throughput figures should be treated as conservative lower bounds;
bare-metal characterization remains future work (\S\ref{sec:future:baremetal}).

These results demonstrate architectural feasibility under controlled replay conditions rather
than universal detection capability. The evaluation covers a single botnet scenario from 2011
with synthetic-trained classifiers; generalizability to modern ransomware, DDoS variants, and
encrypted C2 traffic has not been empirically established and is discussed in
Section~\ref{sec:limitations}. Behavioral indicators associated with ransomware propagation
and lateral-movement patterns are detected in the Neris scenario; direct evaluation against
post-2020 ransomware families remains Future Work~\S\ref{sec:future:corpus}.

All experiments were conducted in a reproducible Vagrant/VirtualBox environment running on a
consumer laptop, ensuring result accessibility and independent reproducibility.

This work was developed through a structured multi-model AI collaboration methodology ---
the \emph{Consejo de Sabios} (Council of Wise Men), a multi-LLM peer review methodology ---
documented in Section~\ref{sec:consejo}. ML Defender is released under the MIT license.
\end{abstract}

\tableofcontents
\newpage

% ============================================================
\section{Introduction}
% ============================================================

The threat is not abstract. On March 5, 2023, the Hospital Cl\'inic de Barcelona suffered a
ransomware attack that paralyzed its emergency services, laboratory, and pharmacy, encrypted
4.5~TB of patient data, and prompted a ransom demand of 4.5~million USD --- which the
institution refused to pay~\cite{incibe2023}. That same day, the author of this work was
leaving a regional hospital in Extremadura after receiving treatment for porphyria --- a rare
metabolic disease that his local hospital had diagnosed and managed. Watching the news, the
thought was immediate and personal: \emph{if it can happen to a major teaching hospital, it
can happen to mine.}

This was not the first time that the author had witnessed the human cost of ransomware
firsthand. Years earlier, the collapse of a close friend's small business had been caused by
an encryption attack that locked every document and file in his office. The author lacked the
technical skills to help. That failure left a debt.

These two events --- the friend's destroyed business, the hospital on television ---
converged into a single design constraint: the solution had to be open-source, it had to run
on commodity hardware, and it had to act in real time. Not forensics. Not post-mortem
analysis. Interception.

The analogy that emerged from this motivation is biological. Network traffic behaves like a
circulatory system: datagrams are the bloodstream, and malicious flows are pathogens. A
useful defense must classify traffic continuously and in real time --- capturing the source IP
of the attacker and blocking it at the firewall before damage propagates. This framing shaped
every architectural decision in ML Defender.

This work therefore addresses the following research question: \emph{can an embedded
machine-learning network detection and response system running entirely on commodity hardware
provide meaningful real-time protection for organizations lacking enterprise security
infrastructure?}

\paragraph{A note on terminology.}
ML Defender (aRGus NDR) implements \emph{Network Detection and Response} (NDR) capabilities:
real-time capture, ML-based classification, and automated blocking at the network layer.
Future work toward full \emph{Endpoint Detection and Response} (EDR) functionality via a
planned lightweight endpoint agent (\S\ref{sec:future}, FEAT-EDR-1) will be reflected in a
corresponding rename when that capability is delivered.

\paragraph{The gap.}
Existing open-source NIDS --- most notably Snort and Suricata~\cite{oisf2010} --- operate
primarily on signature matching and rule-based heuristics. While effective against known
attack patterns, they offer limited resilience against novel ransomware variants and volumetric
DDoS campaigns. Machine learning-based NIDS have been extensively studied~\cite{buczak2016,
pinto2023}, but the gap between research prototype and deployable system remains wide: most
published systems are evaluated on benchmark datasets without accompanying implementation,
require GPU infrastructure, or assume dedicated security operations teams. Healthcare retained
its position as the most targeted sector in 2025, accounting for 22\% of disclosed ransomware
attacks, with an average breach cost of 7.42M~USD~\cite{blackfog2025,ibmsecurity2025}.

\paragraph{Our approach.}
ML Defender (aRGus NDR) addresses this gap with three design principles:
(1)~\emph{embedded inference} --- all ML classification runs in-process at sub-microsecond
latency;
(2)~\emph{commodity deployment} --- the full six-component pipeline runs on bare-metal Linux
hardware costing approximately 150--200~USD;
(3)~\emph{active response} --- upon detection, the attacker's IP is blocked immediately at
the firewall via ipset/iptables.

Beyond the technical system, this paper documents a second contribution: the \emph{Consejo
de Sabios} methodology for human-AI collaborative engineering.

% ============================================================
\section{Background and Related Work}
% ============================================================

\paragraph{Signature-based NIDS.}
Snort~\cite{roesch1999} and Suricata~\cite{oisf2010} apply rule-based pattern matching
against packet payloads and flow metadata. Their fundamental limitation is reactive:
detection requires a prior known signature, a window that historically ranges from hours to
weeks.

\paragraph{ML-based NIDS.}
Anderson \& McGrew~\cite{anderson2016} demonstrated the feasibility of contextual flow data
for encrypted malware classification. Mirsky et al.~\cite{mirsky2018} proposed Kitsune, an
ensemble of autoencoders for unsupervised anomaly detection operating on flow statistics. The
CTU-13 dataset~\cite{garcia2014}, which ML Defender uses for evaluation, has become a
standard benchmark for botnet traffic detection. Despite this body of work, a persistent gap
exists between academic prototype and production system.

\paragraph{Flow-based ML NIDS.}
ML Defender operates exclusively at the flow level, extracting 28 behavioral features per
flow without examining payload content --- ensuring that payload encryption does not
compromise detection capability.

\paragraph{eBPF and XDP-based network monitoring.}
Recent work has explored eBPF/XDP as a foundation for high-performance network security.
ML Defender follows the XDP-capture-plus-userspace-inference architecture: XDP provides
zero-copy packet capture before the kernel networking stack, while all ML inference executes
in userspace at sub-microsecond latency. To our knowledge, open-source projects combining
eBPF capture with userspace ML inference remain rare; most production-grade alternatives
either offload inference to the cloud or require GPU hardware.

\paragraph{Embedded inference.}
Commercial EDR vendors (CrowdStrike, SentinelOne, Carbon Black) implement proprietary
embedded ML pipelines, but neither their architectures nor their training data are publicly
available. ML Defender demonstrates that embedded Random Forest inference achieves
sub-microsecond latency without requiring GPU acceleration or cloud offload.

\paragraph{Human-AI collaborative engineering.}
Structured methodologies for multi-model peer review have not, to our knowledge, been
formally documented in the systems security literature. Section~\ref{sec:consejo} describes
the \emph{Consejo de Sabios} methodology.

% ============================================================
\section{Threat Model}
% ============================================================

\subsection{Adversary Capabilities}
The system assumes an adversary capable of launching attacks from external hosts or
compromised internal machines, including: botnet C2 communication, DDoS (including
amplification attacks), ransomware propagation via SMB lateral movement, and network
reconnaissance.

\subsection{Adversary Limitations}
The system assumes the attacker cannot: fully mimic benign traffic distributions across all
monitored features simultaneously; compromise the ML Defender host; manipulate feature
extraction; or tamper with HMAC-SHA256 protected event logs.

\subsection{Detection Surface}
Detection occurs at the network flow level. ML Defender focuses on connection frequency,
port diversity, TCP flag distributions, packet inter-arrival timing, flow burst behavior, and
destination diversity --- features detectable even when payloads are encrypted.

\subsection{Out-of-Scope Attacks}
Application-layer attacks generating statistically normal traffic; low-and-slow attacks
within benign statistical distributions; insider threats; attacks targeting the ML Defender
host itself. Future work in Section~\ref{sec:future} addresses some of these through expanded
detection vectors (FEAT-NET-1, FEAT-AUTH-1).

% ============================================================
\section{Architecture}
% ============================================================

\subsection{Pipeline Overview}

Six components connected via ZeroMQ and Protocol Buffers:

\begin{enumerate}
  \item \textbf{eBPF/XDP Sniffer} --- zero-copy packet capture at the XDP hook before kernel
    networking stack.
  \item \textbf{Ring Consumer} --- reconstructs bidirectional flow state via
    ShardedFlowManager, computes 28 of 40 planned features. Eleven features use
    \texttt{MISSING\_FEATURE\_SENTINEL = -9999.0f}.
  \item \textbf{Fast Detector} --- rule-based heuristic engine. FPR on purely benign traffic:
    6.61\% (\S\ref{sec:eval:ablation}).
  \item \textbf{ML Detector} --- embedded Random Forest across four threat classes. Three
    classifiers (Ransomware, DDoS, Traffic) in pure C++20; Internal threats via ONNX Runtime.
    Latency: 0.24--1.06~$\mu$s.
  \item \textbf{RAG Subsystem} (\texttt{rag-ingester} + \texttt{rag-security}) ---
    rag-ingester ingests detection events into FAISS + SQLite. rag-security exposes a natural
    language query interface via local TinyLlama --- deliberately local to avoid exfiltrating
    live network metadata to cloud-hosted LLMs.
  \item \textbf{Firewall ACL Agent} --- applies ipset/iptables rules on confirmed detection
    events.
\end{enumerate}

\subsection{Three Functions of the Engine}

\emph{Function 1 --- Real-time detection, classification, and response.}
End-to-end from packet capture to firewall block in milliseconds.

\emph{Function 2 --- Reliable dataset generation.}
ML Detector and Firewall ACL Agent produce structured HMAC-verified CSV logs continuously,
forming the basis for retraining future ensemble models.

\emph{Function 3 --- Real-time observability (the digital microscope).}
The RAG subsystem provides a semantic query interface over the live event stream.

\subsection{Dual-Score Detection}

\begin{equation}
  score_{\text{final}} = \max\!\left(score_{\text{fast}},\; score_{\text{ml}}\right)
  \label{eq:dualscore}
\end{equation}

This is the \emph{arithmetic maximum} over two continuous scores in $[0,1]$, not a logical
OR over binary decisions. Both detectors output continuous confidence scores; the maximum
operator selects the higher-confidence assessment. The ML Detector's primary role is false
positive suppression: the Fast Detector generates $\text{FPR} = 6.61\%$ on purely benign
traffic; the ML Detector reduces real production blocks to zero on the same corpus
(approximately 500-fold reduction). Alternative AND-based consensus policies are planned
(ADR-007, \S\ref{sec:future:consensus}).

\subsection{Transport and Serialization}

ZeroMQ for asynchronous message passing; Protocol Buffers (proto3) for serialization; etcd
for configuration distribution; ChaCha20-Poly1305 for inter-component encryption (selected
for performance on hardware without AES-NI, including ARMv8); HMAC-SHA256 for CSV log
integrity.

\subsection{Deployment Model}

Bare-metal Linux (kernel $\geq 5.8$), dual NIC. Validated in Vagrant/VirtualBox. Production
target: ${\sim}150$--200~USD commodity hardware.

\subsection{Service Coordination (etcd-server)}

Fulfills component registration, JSON config distribution, and ChaCha20 seed provision.
ChaCha20 seed distribution via etcd is \textbf{not recommended for production use} ---
peer-to-peer seed negotiation is under design. etcd is currently single-node; HA etcd
planned.

\subsection{Fast Detector Heuristics (DEBT-FD-001)}

Four heuristics in a 10-second sliding window:

\begin{itemize}
  \item External IP velocity: \texttt{THRESHOLD\_IP\_VELOCITY = 5}
  \item SMB connection count: \texttt{THRESHOLD\_SMB\_CONNS = 3}
  \item Port diversity: \texttt{THRESHOLD\_PORT\_SCAN = 10}
  \item TCP RST ratio: \texttt{THRESHOLD\_RST\_RATIO = 0.20}
\end{itemize}

All thresholds are compile-time \texttt{constexpr} constants --- DEBT-FD-001, scheduled for
JSON migration in PHASE 2 (ADR-006).

% ============================================================
\section{Implementation}
% ============================================================

\subsection{Language and Build System}
C++20. CMake with four build profiles: production, debug, TSAN, ASAN.

\subsection{eBPF/XDP Packet Capture}
XDP hook attachment before SKB allocation. Zero-copy. All parameters loaded from
\texttt{sniffer/config/sniffer.json} at startup.

\subsection{Feature Extraction}

ShardedFlowManager partitions flow tracking across shards, eliminating the ISSUE-003
thread-isolation bug (89 of 142 features silently lost in the original single-threaded
FlowManager). 28 of 40 features computed; 11 use
\texttt{MISSING\_FEATURE\_SENTINEL = -9999.0f}.

\paragraph{Sentinel value selection rationale.}
Split-domain analysis confirmed all Random Forest thresholds lie within $[0.0, 5.1]$.
$-9999.0\text{f}$ lies entirely outside this domain, routing deterministically to the left
child of every split --- non-informative, not influencing classification.
\texttt{quiet\_NaN()} was rejected (undefined comparison behavior across compilers);
$0.5\text{f}$ was rejected (within domain, spurious classification influence). One feature
uses $0.5\text{f}$ as a semantic sentinel for TCP half-open state (deliberate,
within-domain meaningful default).

\subsection{Embedded Random Forest and Synthetic Training Data}

Three classifiers (Ransomware, DDoS, Traffic) compiled as C++20 data structures. Internal
threats classifier via ONNX Runtime --- deliberate to measure latency differential and
footprint trade-off. Models trained with scikit-learn, transpiled to C++20 via custom script;
zero Python or external runtime dependency.

Training/evaluation separation: all classifiers trained on synthetic data; CTU-13 Neris held
out entirely for evaluation.

\paragraph{Synthetic Dataset Methodology.}
Informed by CTU-13 (flow statistics), CIC-IDS2017 (protocol ratios), and MAWI (benign
traffic). Four bias mitigation strategies: IP Space Partitioning (10.0.0.0/8 training vs
147.32.0.0/16 evaluation), Temporal Decoupling, Protocol Ratio Calibration, Feature Boundary
Validation.

\paragraph{Known Limitations.}
Reflects attack patterns circa 2011--2017. Temporal dynamics of real C2 channels simplified
to fixed-window aggregates. Zero-day variants may evade classification.

\subsection{Cryptographic Transport}
ChaCha20-Poly1305 + LZ4 + HMAC-SHA256. Crypto-transport layer extracted as an independent
library.

\subsection{Trace Correlation}

\texttt{trace\_id = SHA-256(src\_ip, dst\_ip, canonical\_attack\_type, timestamp\_bucket)}.
TraceIdPolicy: ransomware=60s, DDoS=10s, SSH brute=30s. Validated: 46/46 unit tests passing.

\subsection{Testing and Validation}

70 tests total: crypto (3), etcd-hmac (12), ml-detector (9), trace\_id (46). 100\% pass
rate. F1 ground truth persisted to \texttt{docs/experiments/f1\_replay\_log.csv}.

\subsection{Development Environment}

Vagrant/VirtualBox. Host: MacBook Pro M2 Pro 16~GB. Guests: \texttt{defender} (Ubuntu 24.04
LTS, Linux 6.x, dual vNIC) + \texttt{client} (traffic generator).

% ============================================================
\section{The Consejo de Sabios: A Methodology for Human-AI Collaborative Engineering}
\label{sec:consejo}
% ============================================================

\subsection{Motivation}

The development of ML Defender spans kernel programming (eBPF/XDP), distributed systems
(ZeroMQ, etcd), cryptography, machine learning, systems security, and academic writing. The
author operated as an independent researcher without institutional affiliation, without a
team, and without access to a formal peer review process. The solution was to construct one.

\subsection{The Methodology}

Seven large language models served as co-reviewers across all development phases:
\textbf{Claude} (Anthropic), \textbf{Grok} (xAI), \textbf{ChatGPT} (OpenAI),
\textbf{DeepSeek}, \textbf{Qwen} (Alibaba), \textbf{Gemini} (Google), \textbf{Parallel.ai}.

These models were engaged as intellectual peers --- presented with architectural proposals,
asked to identify failure modes, invited to challenge assumptions. The methodology operates
on a principle borrowed from ensemble learning: \emph{a diverse ensemble of independent
evaluators produces more robust judgments than any single evaluator, however capable.}

\subsection{How It Worked in Practice}

Convergent recommendations across models increased confidence. Divergent recommendations
triggered deeper investigation. Decisions such as the $-9999.0\text{f}$ sentinel value, the
trace\_id correlation architecture, and the ShardedFlowManager design emerged from or were
refined through these dialogues.

\subsection{Test Driven Hardening}

TDH methodology: (1)~Witness --- characterize anomalous behavior. (2)~Demonstration test ---
produce a minimal test that fails and exposes the mechanism. (3)~Independent solution ---
each model proposes a fix against the shared contract. (4)~Human arbitration --- author
selects, weighing correctness, maintainability, and coherence.

TDH differs from TDD: the test is written \emph{before the fix}, as a proof of failure ---
not before the feature, as a specification.

\subsection{Attribution}

The author remained the final arbiter of all decisions; all empirical validation was
performed by the human author. The seven models are credited as intellectual co-contributors
in the project's commit history, ADRs, and this paper. In the context of arXiv submission,
the models are acknowledged in \S\ref{sec:acknowledgments}; formal co-authorship conventions
for AI systems remain an evolving area, and this paper adopts the most transparent available
approach consistent with emerging ACM and IEEE norms on AI contribution disclosure.

\subsection{Why This Matters}

The Consejo de Sabios democratizes access to deep technical peer review for independent
researchers without institutional affiliation --- parallel to ML Defender's mission of
democratizing enterprise-grade security for organizations without enterprise budgets.

\subsection{Limitations}

Models cannot run code or observe system behavior directly. Models can hallucinate --- several
incorrect technical suggestions were identified and rejected. The methodology augments human
judgment; it does not replace it.

% ============================================================
\section{Formal System Model}
% ============================================================

\subsection{Network Traffic Representation}

Packets $P = \{p_1, \dots, p_n\}$ grouped into bidirectional flows by canonical five-tuple
$(src, dst, sport, dport, proto)$ via ShardedFlowManager $\Phi : P \rightarrow F$. Flows
are bidirectional five-tuple aggregates; behavioral features are computed over a 10-second
sliding window as configured in \texttt{sniffer.json}. Flows are closed after an inactivity
timeout as defined by the sliding window expiry.

\subsection{Feature Extraction}

\begin{equation}
  x_f = g(P_f, W), \quad x_f \in \mathbb{R}^{28}
\end{equation}

\subsection{Fast Detector}

\begin{equation}
  H(x_f) = \max_j h_j(x_f), \quad h_j(x_f) \in \{0, 1\}
\end{equation}

\subsection{Machine Learning Detector}

\begin{equation}
  M(x_f) = \frac{1}{T} \sum_{k=1}^{T} T_k(x_f)
\end{equation}

\subsection{Detection Policy}

\begin{equation}
  S(x_f) = \max\!\left(H(x_f),\; M(x_f)\right) \geq \theta
\end{equation}

Thresholds: $\theta_{\text{ransomware}} = 0.85$, $\theta_{\text{DDoS}} = 0.90$,
$\theta_{\text{traffic}} = 0.80$, $\theta_{\text{internal}} = 0.85$.

\subsection{Response Function}

\begin{equation}
  R(f) = \begin{cases}
    \texttt{block}(src_f) & \text{if } S(x_f) \geq \theta \\
    \texttt{allow}        & \text{otherwise}
  \end{cases}
\end{equation}

Blocking is applied per source IP. Blocks are currently persistent for the session duration;
TTL-bounded blocks with configurable expiry are Future Work (\S\ref{sec:future:response}).

\subsection{System Determinism}

Given identical packet input $P$, the pipeline produces identical outputs. Determinism is a
prerequisite for the reproducibility guarantees described in Section~\ref{sec:reproducibility}.

% ============================================================
\section{Evaluation}
\label{sec:eval}
% ============================================================

\subsection{Experimental Setup}

\begin{table}[h]
\centering
\caption{Experimental hardware and software specifications.}
\label{tab:setup}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Host machine       & Apple MacBook Pro (Apple Silicon M2 Pro) \\
Host RAM           & 16 GB \\
Guest OS           & Ubuntu 24.04 LTS (Linux kernel 6.x) \\
Virtualization     & VirtualBox 7.2, bridged NIC mode \\
VM vCPUs           & 4 (estimated ${\sim}$2.4 GHz under VirtualBox) \\
VM RAM             & 4 GB (8 GB allocated for stress tests) \\
Network adapter    & Paravirtualized (VirtIO) \\
Virt.\ overhead    & Estimated 5--15\% vs bare-metal \\
Pipeline config.   & Full 6-component, dual-VM \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Methodological note on metrics.}
\texttt{calculate\_f1\_neris.py} measures \textbf{Fast Detector} alerts
(\texttt{[FAST ALERT]} in sniffer.log) against CTU-13 Neris ground truth IPs. The ML
Detector's \texttt{attacks=N} statistic reports flows exceeding production thresholds and
triggering firewall blocks. Both are reported; F1=0.9985 is the primary system detection
metric.

\subsection{Dataset}

\paragraph{CTU-13 Neris.}
19,135 total flows; \textbf{646 malicious flows} (host 147.32.84.165) constitute the TP
ground truth. Captured 2011. Garcia et al.~\cite{garcia2014} document that the Neris botnet
exhibits C2 communication patterns concentrated in a behavioral subset --- specifically,
flows matching IRC-based C2 channel signatures (connection bursts, SMB lateral movement, DNS
query anomalies). The remaining flows represent background traffic on the infected host that
do not exhibit malicious behavioral signatures and are therefore not counted as ground-truth
positives. This distinction aligns with standard practice in NIDS
evaluation~\cite{sharafaldin2018,mirsky2018}.

\paragraph{BigFlows.}
A CTU-13 derived dataset containing 40,467 flows and 791,615 packets, used for high-throughput
replay experiments. 38,064 flows were processed by ml-detector. Treated conservatively as
\emph{probable benign} --- no official ground truth exists for this CTU-13 scenario.

\paragraph{Synthetic training data.}
All classifiers trained exclusively on synthetic data; CTU-13 Neris held out entirely for
evaluation.

\subsection{Detection Performance}

\begin{table}[h]
\centering
\caption{Detection performance --- Fast Detector pipeline on CTU-13 Neris.}
\label{tab:detection}
\begin{tabular}{ll}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
F1 Score                       & \textbf{0.9985} \\
Precision                      & 0.9969 \\
Recall                         & 1.0000 \\
True Positives                 & 646 \\
False Positives                & 2 \\
False Negatives                & 0 \\
True Negatives                 & 12,075 \\
Total events processed         & 12,723 \\
FPR (Neris evaluation)         & \textbf{0.0002\%} \\
ML Detector high-conf.\ attacks & 12 \\
FPR Fast Detector (bigFlows)   & 6.61\% \\
FP reduction (Fast $\to$ ML)   & \textbf{${\approx}$500-fold} \\
\bottomrule
\end{tabular}
\end{table}

F1=0.9985 stable across replay runs at total\_events = 12,563 / 12,605 / 12,723 / 13,930.

\subsection{False Positive Identification and Confusion Matrix}

\begin{table}[h]
\centering
\caption{False positive identification.}
\label{tab:fps}
\begin{tabular}{llll}
\toprule
\textbf{FP} & \textbf{Source} & \textbf{Destination} & \textbf{Classification} \\
\midrule
FP-1 & 192.168.56.1 & 224.0.0.251     & mDNS multicast (VirtualBox host-only) \\
FP-2 & 192.168.56.1 & 192.168.56.255  & Broadcast (VirtualBox host-only) \\
\bottomrule
\end{tabular}
\end{table}

Both are VirtualBox host-only adapter artifacts --- absent in bare-metal deployments.

\paragraph{Confusion matrix.}
\[
  \text{Precision} = \frac{646}{648} = 0.9969, \quad
  \text{Recall}    = \frac{646}{646} = 1.0000, \quad
  \text{F1}        = 0.9985
\]

\subsection{Inference Latency}

\begin{table}[h]
\centering
\caption{Per-class inference latency.}
\label{tab:latency}
\begin{tabular}{lll}
\toprule
\textbf{Classifier} & \textbf{Implementation} & \textbf{Latency} \\
\midrule
DDoS        & Embedded C++20 & 0.24~$\mu$s \\
Traffic     & Embedded C++20 & 0.37~$\mu$s \\
Internal    & ONNX Runtime   & 0.33~$\mu$s \\
Ransomware  & Embedded C++20 & 1.06~$\mu$s \\
\bottomrule
\end{tabular}
\end{table}

Latency values are mean inference times measured via
\texttt{std::chrono::high\_resolution\_clock} around the \texttt{classify()} call within the
ml-detector process, excluding inter-component ZeroMQ transport latency, over $N \geq 1{,}000$
inference calls per classifier during the CTU-13 Neris replay.

\subsection{Ablation Study}
\label{sec:eval:ablation}

\textbf{Config A --- Fast Detector Only:} 2,517 FP on 38,064 bigFlows (FPR=6.61\%),
Recall=1.0000 on Neris.

\textbf{Config B --- ML Detector Only (inferred):} 5 detections on 38,064 bigFlows, all
below production thresholds (max confidence 68.97\%) --- 0 real blocks. Neris: attacks=12.

\textbf{Config C --- Dual (deployed):} F1=0.9985, FPR=0.0002\%. Only Config C fully
validated end-to-end.

\subsection{Interpretation and Comparison with State of the Art}

F1=0.9985 and Recall=1.0000 reflect the statistical separability of the Neris botnet
scenario, not a claim about arbitrary threats. The synthetic training successfully captured
the Neris behavioral signature; generalization to other families is an open empirical
question.

\begin{table}[h]
\centering
\footnotesize
\caption{Indicative comparison with representative NIDS approaches. \emph{Caution:
cross-paper comparison is indicative only --- systems evaluated on different datasets, at
different operating points, and under different experimental conditions. Direct experimental
comparison was not performed.}}
\label{tab:comparison}
\begin{tabular}{lllllll}
\toprule
\textbf{System} & \textbf{Dataset} & \textbf{F1} & \textbf{Latency} & \textbf{HW} & \textbf{Block} & \textbf{Notes} \\
\midrule
\textbf{ML Defender} & CTU-13 Neris & \textbf{0.9985} & 0.24--1.06~$\mu$s & x86/ARM & Yes & Synthetic, dual-score \\
Kitsune~\cite{mirsky2018} & Custom & ${\sim}$0.982 & ${\sim}$50~$\mu$s & x86 & No & Autoencoder ensemble \\
Suricata (rules) & Various & ${\sim}$0.891 & ${\sim}$10~$\mu$s & x86 & Yes & Signature+heuristics \\
RF~\cite{sharafaldin2018} & CIC-IDS2017 & ${\sim}$0.943 & ${\sim}$200~$\mu$s & x86 & No & Batch, not embedded \\
Snort & N/A & N/A & 1--5~ms & x86 & Yes & Rule-based, no ML \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ransomware Detection: Scope and Validation}

\paragraph{What is validated.}
Behavioral indicators associated with ransomware propagation and lateral-movement patterns
are detected in the 2011 Neris botnet scenario: SMB lateral movement (ports 445/139 burst),
high-rate connection attempts, DNS query anomalies.

\paragraph{What is not validated.}
Direct evaluation against post-2020 ransomware families (LockBit, BlackCat/ALPHV, Cl0p) has
not been performed. Modern ransomware uses encrypted C2, adaptive beaconing with jitter,
staged delivery --- potentially different statistical signatures.

\paragraph{Future work.}
Direct evaluation against modern ransomware captures is the highest-priority evaluation
expansion (\S\ref{sec:future:corpus}), with Sebastian Garcia (CTU Prague) as potential
collaborator.

\subsection{Throughput Stress Test and Resource Utilization}

\begin{table}[h]
\centering
\caption{Stress test replay results --- progressive Mbps escalation (DAY 87).}
\label{tab:stress}
\begin{tabular}{llllllll}
\toprule
\textbf{Run} & \textbf{Dataset} & \textbf{Req.} & \textbf{Actual} & \textbf{PPS} & \textbf{Packets} & \textbf{Failed} & \textbf{Errors} \\
\midrule
1 & Neris    & 10 Mbps  & 9.60 Mbps  & ${\sim}$3,700   & 320,524   & 2,630$^{\dagger}$ & 0 \\
2 & Neris    & 25 Mbps  & 11.28 Mbps & ${\sim}$4,300   & 320,524   & 2,630$^{\dagger}$ & 0 \\
3 & bigFlows & 50 Mbps  & 34.68 Mbps & 9--10K          & 791,615   & 0                 & 0 \\
4 & bigFlows & 100 Mbps ($\times$3) & 33.16 Mbps & 9--11K & 2,374,845 & 0           & 0 \\
\bottomrule
\end{tabular}
\medskip
\footnotesize{$\dagger$ Pre-existing pcap artifact --- identical count at 10 and 25 Mbps,
confirming origin in pcap structure, not pipeline saturation.}
\end{table}

\begin{table}[h]
\centering
\caption{Resource utilization during 100 Mbps bigFlows replay (DAY 87).}
\label{tab:resources}
\begin{tabular}{lllll}
\toprule
\textbf{Component} & \textbf{CPU\%} & \textbf{RES} & \textbf{MEM\%} & \textbf{Notes} \\
\midrule
ml-detector       & ${\sim}$315--320\% & 164 MB    & 2.0\%  & 3.2 cores; multi-threaded \\
sniffer           & ${\sim}$88--108\%  & 55--57 MB & 0.7\%  & XDP + ring consumer \\
firewall-acl      & ${\sim}$10--16\%   & 101 MB    & 1.2\%  & Stable throughout \\
rag-ingester      & ${\sim}$12--20\%   & 36--39 MB & 0.5\%  & FAISS indexing \\
etcd-server       & ${\sim}$0.2\%      & 10 MB     & 0.1\%  & Minimal \\
rag-security      & idle               & 1.2 GB VIRT & 15.0\%$^{\ddagger}$ & TinyLlama loaded \\
\midrule
\textbf{Total}    & ${\sim}$65--73\%   & ${\sim}$1,278 MB & --- & ${\sim}$23--30\% idle \\
\bottomrule
\end{tabular}
\medskip
\footnotesize{$\ddagger$ rag-security's 1.2 GB virtual footprint is the TinyLlama model
mapped at startup.}
\end{table}

The throughput ceiling is VirtualBox NIC emulation (${\sim}$33--38 Mbps), not pipeline
logic. Zero packet drops, zero deserialization errors across all 2,374,845 packets. Total
pipeline RAM stable at ${\sim}$1.28~GB (${\pm}18$~MB drift) throughout the 8-minute test.

% ============================================================
\section{Performance Model and Throughput Analysis}
% ============================================================

\subsection{Pipeline Processing Model}

Six stages operating concurrently via ZeroMQ; effective throughput bounded by the slowest
stage:
\begin{equation}
  T_{\max} = \frac{1}{\max_i(L_i)}
\end{equation}

\subsection{Inference Cost}

$\mathcal{O}(T \cdot D)$ --- simple threshold comparisons on compact C++20 data structures
fitting in CPU cache.

\subsection{Throughput: Theoretical Bound and Measured Ceiling}

\paragraph{Theoretical upper bound (single-core, ML inference only):}
\begin{equation}
  T_{\text{theory}} \approx \frac{1}{0.5\;\mu\text{s}} = 2 \times 10^6\;\text{flows/sec}
\end{equation}

\paragraph{Measured virtualized ceiling (DAY 87).}
The VirtIO NIC limits delivery to ${\sim}$33--38~Mbps (${\sim}$9,000--11,000~pps).
At this ceiling: ml-detector sustains ${\sim}$3.2 cores; total pipeline RAM stable at
${\sim}$1.28~GB; system CPU ${\sim}$65--73\% user, ${\sim}$23--30\% idle.

\subsection{Queue Stability}

Stability requires $\lambda < \mu$. Post-replay drain behavior provides direct empirical
evidence: sniffer CPU drops to near-zero while ml-detector continues draining at full
throughput for several minutes --- confirming the ZeroMQ buffer absorbed the burst cleanly.
\texttt{errors=(deser:0, feat:0, inf:0)} confirmed at all observation points.

\subsection{Scalability}

\textbf{Vertical:} additional CPU cores $\to$ additional parallel flow-processing threads,
linear throughput improvement up to capture bottleneck.

\textbf{Horizontal:} multiple ML Defender nodes across network segments, each processing
local traffic independently while contributing to distributed dataset generation.

% ============================================================
\section{Limitations}
\label{sec:limitations}
% ============================================================

\textbf{10.1 Evaluation Scope.} Single botnet scenario (CTU-13 Neris, 2011). Modern
ransomware communication patterns may differ substantially. Generalizability not empirically
established.

\textbf{10.2 Incomplete Feature Set.} 28 of 40 features implemented; 11 use $-9999.0$f
sentinel. Impact of missing 12 features on broader traffic distributions unknown.

\textbf{10.3 Synthetic Training Data.} Reflects patterns circa 2011--2017. Zero-day
variants may evade classification.

\textbf{10.4 Fast Detector False Positive Rate.} FPR=6.61\% on bigFlows under current
JSON-configurable thresholds (Path B, active since DAY 80). An earlier compile-time baseline
(Path A, DAY 13--79, DEBT-FD-001) produced FPR=76.8\% on the same corpus; DAY 80 JSON
threshold migration achieved a 12$\times$ improvement.

\textbf{10.5 Cryptographic Seed Distribution.} ChaCha20 seed via etcd not recommended for
production. Peer-to-peer negotiation under design.

\textbf{10.6 Single-Node Deployment.} Distributed deployment not validated
experimentally.

\textbf{10.7 etcd High Availability.} Single-node; failure disrupts registration and config
distribution. HA planned.

\textbf{10.8 RAG Index Capacity.} FAISS event store finite capacity not mathematically
characterized.

\textbf{10.9 Virtualization Overhead.} All latency and throughput measurements obtained
under VirtualBox. Reported latencies are conservative upper bounds; reported throughput
(${\sim}$33--38~Mbps) is a conservative lower bound imposed by NIC emulation. Bare-metal
characterization remains future work (\S\ref{sec:future:baremetal}).

\textbf{10.10 BigFlows Ground Truth.} Treated as \emph{probable benign}, not
\emph{confirmed benign}. FPR=0.0002\% is a lower bound.

% ============================================================
\section{Future Work}
\label{sec:future}
% ============================================================

\textbf{11.1 Evaluation Corpus Expansion (P0).}
\label{sec:future:corpus}
CTU-13 additional scenarios, modern ransomware captures, volumetric DDoS datasets. Contact
initiated with Sebastian Garcia (CTU Prague) as potential arXiv endorser and collaborator.

\textbf{11.2 Feature Completion and Importance Analysis.}
Complete remaining 12 real features. Feature importance analysis (Gini impurity reduction
per feature from embedded RF).

\textbf{11.3 Fast Detector JSON Migration (DEBT-FD-001).} Scheduled for PHASE 2.

\textbf{11.4 Consensus Detection Policy (ADR-007).}
\label{sec:future:consensus}
AND-based policy requiring simultaneous agreement before autonomous blocking ---
JSON-configurable for operator selection.

\textbf{11.5 Response Strategy Expansion.}
\label{sec:future:response}
TTL-bounded blocks, traffic quarantine, honeypot redirection.

\textbf{11.6 Cryptographic Hardening.} Peer-to-peer ChaCha20 seed negotiation; etcd HA.

\textbf{11.7 Runtime Reconfiguration (Hot-Reload).} JSON parameter updates without restart.

\textbf{11.8 Graph Attack Engine (ENT-2).} Live attack graphs from CSV event data; STIX 2.1
export.

\textbf{11.9 Fleet-Wide Observability.} Fine-tuned LLM over anonymized aggregated data from
all deployed nodes.

\textbf{11.10 Test Driven Hardening as Independent Research.} TDH generalizability, scaling
with model capability, and measurable outcomes vs single-model consultation.

\textbf{11.11 Bare-Metal Throughput Stress Testing (P1).}
\label{sec:future:baremetal}
Progressive tcpreplay at 100/250/500/1000~Mbps on physical NIC; CPU/memory/packet-drop
monitoring per component. All throughput figures in this paper should be treated as
conservative lower bounds.

\textbf{11.12 Infiltration Vector Coverage.}
ML Defender currently detects ransomware \emph{propagation}. Future work addresses
\emph{infiltration}: FEAT-NET-1 (DNS anomaly/DGA detection), FEAT-NET-2 (threat intelligence
feeds), FEAT-AUTH-1 (authentication log ingestion), FEAT-AUTH-2 (behavioral auth anomaly),
FEAT-EDR-1 (lightweight endpoint agent).

\textbf{11.13 Enterprise Features.} ENT-1 (Federated Threat Intelligence), ENT-2 (Attack
Graph Generation), ENT-3 (P2P Seed Distribution), ENT-4 (Hot-Reload). MIT-licensed core
remains free.

\textbf{11.14 Ethical Deployment in Healthcare Environments.} Not certified for FDA 21 CFR
Part 820 or EU MDR 2017/745. Phased deployment: (1) 30-day monitoring mode with human
approval; (2) TTL-bounded blocks with SOC notification; (3) full autonomous blocking after
site-specific FPR validation.

% ============================================================
\section{Conclusion}
% ============================================================

This paper presented ML Defender (aRGus NDR), an open-source network detection and response
system built in C++20, designed from first principles to protect the organizations that need
protection most and can afford it least.

Evaluated against real botnet traffic from the CTU-13 Neris dataset --- traffic the system
had never seen during training --- ML Defender achieves $\text{F1} = 0.9985$,
$\text{Precision} = 0.9969$, and $\text{Recall} = 1.0000$. The two false positives are fully
identified as VirtualBox network artifacts that do not occur in bare-metal deployments. The
Fast Detector alone produces a 6.61\% FPR on purely benign traffic; the ML Detector reduces
this to zero real production blocks on the same corpus. End-to-end inference latency ranges
from 0.24~$\mu$s to 1.06~$\mu$s per flow, entirely in-process, on commodity hardware costing
approximately 150--200~USD.

Under progressive load testing, the pipeline processes up to ${\sim}$34--38~Mbps --- the
ceiling imposed by VirtualBox NIC emulation --- with zero packet drops, zero errors, and
stable RAM (${\sim}$1.28~GB) across 2.37 million replayed packets. The bottleneck is the
hypervisor, not the software.

These results cover one botnet scenario from 2011. They demonstrate architectural viability
and effective detection within that scope. The limitations in Section~\ref{sec:limitations}
are the honest documentation of a system that does what it claims to do within the boundaries
where it has been tested.

ML Defender was built because a friend lost his business to ransomware, and because a
hospital in Barcelona was paralyzed by an attack on the same day its author was leaving a
regional hospital in Extremadura --- the hospital that had diagnosed his porphyria and saved
his life. The design constraints that followed from those two events were not abstract:
open-source, commodity hardware, real-time response. Not a forensic tool. Not a dashboard. A
system that acts.

The \emph{Consejo de Sabios} represents a new mode of engineering: one that gives a single
independent researcher access to the kind of multi-expert peer review that has historically
required institutional affiliation, research groups, and geographic proximity to centers of
expertise. The \emph{Test Driven Hardening} methodology that emerged from this collaboration
has proven effective against some of the most subtle concurrency and memory bugs in the
codebase.

Both contributions --- the technical system and the collaborative methodology --- are acts of
democratization. The parallel is not accidental. It is the point.

\medskip

\noindent For the hospital in Extremadura. For the friend whose business was broken by
ransomware. For every small organization that has been told, implicitly or explicitly, that
enterprise-grade security is not for them.

\medskip

\noindent \textit{It is now.}

% ============================================================
\section{Reproducibility and Artifact Availability}
\label{sec:reproducibility}
% ============================================================

All experiments are reproducible from the public repository:
\url{https://github.com/alonsoir/argus}

\paragraph{Dataset Setup.}
\begin{lstlisting}[language=bash]
# CTU-13 Neris (scenario 10) -- required for F1 validation
wget https://mcfp.felk.cvut.cz/publicDatasets/CTU-Malware-Capture-Botnet-42/\
botnet-capture-20110810-neris.pcap -O /vagrant/datasets/ctu13/neris.pcap
# Mirror: https://www.stratosphereips.org/datasets-ctu13

# CTU-13 bigFlows -- required for stress test
wget https://mcfp.felk.cvut.cz/publicDatasets/CTU-Malware-Capture-Botnet-42/\
bigFlows.pcap -O /vagrant/datasets/ctu13/bigFlows.pcap
\end{lstlisting}

\paragraph{F1 Evaluation.}
\begin{lstlisting}[language=bash]
make pipeline-stop && make logs-lab-clean && make pipeline-start && sleep 15
make test-replay-neris
python3 scripts/calculate_f1_neris.py \
  /vagrant/logs/lab/sniffer.log --total-events 19135
\end{lstlisting}

\paragraph{Stress Test.}
\begin{lstlisting}[language=bash]
vagrant ssh client -c "sudo tcpreplay -i eth1 --mbps=100 --loop=3 \
  /vagrant/datasets/ctu13/bigFlows.pcap"
\end{lstlisting}

\paragraph{Determinism.}
Classifiers embedded in binary; no stochastic inference components; feature extraction
deterministic. F1=0.9985 confirmed stable at total\_events = 12,563 / 12,605 / 12,723 /
13,930.

% ============================================================
\section*{Acknowledgments}
\label{sec:acknowledgments}
% ============================================================

This work was carried out independently, without institutional funding, without a research
group, and without affiliation to any university or laboratory.

The author engaged seven large language models --- Claude (Anthropic), Grok (xAI), ChatGPT
(OpenAI), DeepSeek, Qwen (Alibaba), Gemini (Google), and Parallel.ai --- as structured
intellectual collaborators across 88 days of continuous development. Their contributions to
architectural decisions, failure mode identification, debugging, test scenarios, and paper
review are documented in the project's commit history and ADRs. The \emph{Test Driven
Hardening} methodology documented in Section~\ref{sec:consejo} emerged directly from this
collaboration.

This paper discloses AI involvement openly and precisely. The author remained the final
arbiter of all decisions; all empirical validation was performed by the human author.

The author wishes to acknowledge the friend whose business was destroyed by ransomware ---
who will recognize himself in these pages. And the Hospital de M\'erida, Extremadura, whose
physicians diagnosed a rare disease and gave the author the years needed to build something
worth building.

\medskip
\noindent This work is dedicated to every hospital, school, and small organization that has
ever been told that enterprise-grade security is beyond their reach.

% ============================================================
% Bibliography
% ============================================================

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}